An elaborate data set on human gait and the effect of mechanical perturbations

### 2014-12-11 paper.tex
Reworked intro with new citations, closes #36
- new bib file (includes some refs I have yet to cite, but maybe should)
- Different intro flavor in light of lack of gait data.

\author[1]{Jason K. Moore}
\author[1]{Sandra K. Hnat}
\author[1]{Antonie van den Bogert}

@@

  Here we share a rich gait data set collected from fifteen subjects walking at
  three speeds on an instrumented treadmill. Each trial consists of 120 seconds
  of normal walking and 480 seconds of walking while being longitudinally
  and/or laterally perturbed during each stance phase with pseudo-random
  fluctuations in the speed of the treadmill belt. A total of approximately 1.5
  hours of normal walking ($>5000$ gait cycles) and 6 hours of perturbed walking
  ($>20,000$ gait cycles) is included in the data set. We provide full body
  marker locations and ground reaction loads in addition to a presentation of
  processed data that includes gait landmarks, 2D joint angles, angular rates,
  and joint torques and the software used for the computations. The protocol is
  described in detail along with additional elaborate meta data. This data can
  likely be useful for validating or generating mathematical models that are
  capable of simulating normal periodic gait and non-periodic, perturbed gaits.

@@

\section*{Introduction}
%
Emerging powered prosthetics are in need of bio-inspired control systems.
Ideally, a powered prosthetic would behave in such a way that the user would
feel like their limb was never disabled. There are a variety of approaches to
developing bio-inspired control systems some of which aim to mimic the
reactions and motion of an able-bodied person. A modern gait lab is able to
collect a variety of kinematic, kinetic, and physiological data from humans
during gait. This data can potentially be used to drive the design of the human
mimicking controller. With a rich enough data set, one may be able to identify
control mechanism used during a human's natural gait and recovery from
perturbations. We have collected data that we believe is richer than previous
gait data sets and may be rich enough for control identification. The data can
also be used for verification purposes for controllers that have been designed
in other manners.

We collected over five and half hours of perturbed gait data from 15 subjects
sampled at 100 hz. The data has been organized and made available for other
research uses.

\section*{Particpants}
%
Fifteen able bodied subjects including four females and eleven males with an
average age of $24\pm4$ years, height of $1.75\pm0.09$ m, mass of $74\pm13$ kg
particpated in the study. The data has been anonymized with respect to the
participants' identities and an unique identification number was assigned to
each subject. A selection of the meta data collected for each subject is shown
in Table~\ref{tab:subjects}.
%
\begin{table*}
  \cprotect\caption{Information about the 15 participants. The final three
    columns give the trial numbers associated with each nominal treadmill
    speed. Additional trials with subject 0 are trials with no subject, i.e.
    unloaded trials that can be used for force compensation purposes, and are
    not shown in the table. Generated by \verb|src/subject_table.py|.}
  \centering
  \input{tables/subjects.tex}
  \label{tab:subjects}
\end{table*}

@@

\begin{itemize}
  \item A ForceLink R-Mill treadmill which has dual 6 degree of freedom force
    plates, independent belts for each foot, and motion capabilities (lateral,
    pitch).
  \item A 10 camera Motion Analysis motion capture system which includes Osprey
    cameras, the Cortex software (3.1.1.1290), and hardware for collecting
    analog and camera data simultaneously.
  \item National Instruments USB-6255 data aquisition unit.
  \item Four Sparkfun ADXL330 Triple Axis Accelerometer Breakout boards
    attached to the treadmill.
  \item Motek Medical’s D-Flow software (3.16.1 to 3.16.2) and visual display
    system.
\end{itemize}

@@

\section*{Protocol}
%
The experimental protocol consisted of static measurements and walking on the
treadmill for 10 minutes under unperturbed and perturbed conditions. Before a
set of trials on the same day the following happened:
%
\begin{itemize}
  \item Calibration of the motion capture system.
  \item Subject changes into atheltic shoes, lyrca shorts, sports bra, baseball
    cap, and rock climbing harness.
  \item All 47 markers are applied directly to the skin except for the heel,
    toe, and head markers.~\footnote{The sacrum and rear pelvic markers were
    placed on the shorts for a small number of the subjects}.
  \item Subjects self-reported age, gender, and mass.
  \item Height is measured by the experimentalist.
  \item Four reference photographs (front, back, right, left) are taken of
    subjects marker locations.
\end{itemize}

After a briefing by the expermentalist on the trial protocol the subject
follows the verbal instructions of the experimentalist and the on-screen
instructions from the video display in front of the subject. The protocal goes
as so:
%
\begin{enumerate}
  \item Subject steps onto the treadmill and markers are identified with
    Cortex.
  \item The safety cable is attached loosely to the rock climbing harness such
    that no undue forces are acting on the subject during walking, but that the
    harness will prevent a full fall.
  \item The subject starts by stepping on sides of treadmill so that feet don't
    touch the force plates and the force plate signals are zero'd.
  \item Once notified by the screen, the subject stands in the initialization
    pose: standing straight up, looking forward, arms out by their sides (~45
    degrees) and the event is manually recorded by the operator.
  \item A countdown to the first normal walking phase is displayed. At the end
    of the countodwn the treadmill starts at the specified speed and the
    subject is instructed to walk normally, to focus on the road in the
    screen, and not to look at their feet.
  \item After 1 minute of normal walking, the longitudinal perturbation phase
    begins.
  \item After 8 minutes of walking under the influence of the perturbations,
    the normal walking phase begins.
  \item After 1 minute, a countdown is shown and the treadmill decelerates to a
    stop.
  \item The subject is instructed to step off of the force plates for 10
    seconds.
  \item The subject could take a break before each additional trial.
\end{enumerate}

@@

The resulting output from the MATLAB script and Simulink produces a comma-
delimited text file of six columns: timestamp, slow, normal, and fast walking
perturbation signals, and slow and fast running signals.  These running signals
were generated in the event the testing protocol expanded from a walking study
to a running study, though they were not used in the data set.  The actual
motion of the treadmill belts are compared to the original input signals as
shown in Figure~\ref{fig:input_output}.   The standard deviations of the
outputs were calculated at 6.07\%, 9.68\%, and 12.68\% for the the 0.8, 1.2,
and 1.6 $m \cdot s^{-1}$ speeds, respectively.

@@

The text file output from the mocap module in DFlow is a tab delimited file.
The first line is the header and contains a time stamp column, frame number
column, marker position columns, force plate force/moment columns, force plate
center of pressure columns, other analog columns, and potentially results from
the real time Human Body Model \cite{Bogert2013} which is included with the
D-Flow software. These are detailed below. The numerical values of the
measurements are provided in decimal floating point notation with 6 decimals of
precision, e.g. 123456.123456 [\%1.6f].

@@

\begin{description}
  \item[TimeStamp] The monotonically increasing computer clock time when D-Flow
    recieves a frame from Cortex. These are approximately at 100 hz and given
    in seconds.
  \item[FrameNumber] Monotonically increasing positive integers that correspond
    to each frame received from Cortex.
  \item[Marker Coordinates] Any column that ends in \verb+.PosX+, \verb+.PosY+,
    or \verb+.PosZ+ are marker coordinates expressed in Cortex's Cartesian
    reference frame. These values are in meters.
  \item[Ground Reaction Loads] There are three ground reaction forces and three
    ground reaction moments recorded by each of the two force plates in Newtons
    and Newton-Meters, respectively. The prefix for these columns is either
    \verb+FP1+ or \verb+FP2+ and represents either force plate 1 (left) or 2
    (right). The suffixes are either \verb+.For[XYZ]+, \verb+.Mom[XYZ]+ for the
    forces and moments, respectively. The force plate voltages are sampled at a
    much higher frequency than the cameras, but delivered at the Cortex camera
    sample rate, 100 Hz through the D-Flow mocap module. A force/moment
    calibration matrix stored in Cortex converts the voltages to forces and
    moments before sending it to D-Flow [1]. Cortex also computes the center of
    pressure from the forces, moments, and force plate dimensions. These have
    the same prefixes for the plate number, have the suffix \verb+.Cop[XYZ]+,
    and are in meters.
  \item[Analog Channels] Several analog signals are recorded under column
    headers \verb+Channel[1-99].Anlg+. These correspond to analog signals
    sampled by Cortex and correspond to the 96 analog-channel National 
    Instruments 6255 DAQ box. The first twelve are the voltages from the 
    force plate load cells. We also record the acceleration of 4 points on the
    treadmill base in channels 61-72.  TvdB: are these useful?  If so, we need to 
    provide information on how to process these channels.  If not, maybe
    not mention them at all?
\end{description}

@@

\begin{description}
  \item[analog-channel-map] A mapping of the strings D-Flow assigns to signals
    emmitted from the analog channels of the NI USB-6255 to names the user
    desires.
  \item[cortex-version] The version of Cortex used to record the trial.
  \item[datetime]A date formatted string giving the date of the trial in the
    \verb|YYYY-MM-DD| format.
  \item[dflow-version] The version of D-Flow used to record the trial.
  \item[events] A key value map which prescribes names to the alphabetic events
    recorded in the record file.
  \item[files] A key value mapping of files associated with this trial
    where the key is the DFlow file type and the value is the path to the file
    relative to the meta file.
  \item[hardware-settings] There are tons of settings for the hardware in both
    D-Flow, Cortex, and the other software in the system. This contains any
    non-default settings.
    \begin{description}
      \item[high-performance] A boolean value indicatin whether the D-Flow high
        performance setting was on (True) or off (False).
    \end{description}
  \item[id] An unique three digit integer identifier for the trial. All of the
    file names and directories associated with this trial include this number.
  \item[marker-map] A key value map which maps marker names in the mocap file
    to the user's desired names for the markers.
  \item[marker-set] Indicates the HBM~\cite{Bogert2013} marker set used during
    the trial, either full, lower, or NA.
  \item[nominal-speed] A float representing the nominal desired treadmill speed
    during the trial.
  \item[nominal-speed-units] (string) The units of the nominal speed.
  \item[notes] Any notes about the trial.
  \item[pitch] A boolean that indicates if the treamdill pitch degree of
    freedom was acutated during the trial.
  \item[stationary-platform] A boolean that indicates whether the treadmill
      sway or pitch motion was actuated during the trial. If this flag is
      false, the measured ground reaction forces must be compensated for the
      inertial affects and be expressed in the motion capture reference frame.
  \item[subject-id] An integer corresponding to the subject in the trial.
  \item[sway] A boolean that indicates if the treadmill lateral (sway) degree
    of freedom was acutated during the trial.
\end{description}

\begin{listing*}
  \small
  \begin{minted}[gobble=4]{yaml}
    study:
       id: 1
       name: Gait Control Identification
       description: Perturb the subject during walking and running.
    subject:
       id: 8
       age: 20
       mass: 70.0
       mass-units: kilograms
       height: 1.572
       height-units: meters
       knee-width-left: 107.43
       knee-width-right: 107.41
       knee-width-units: millimeters
       ankle-width-left: 70.52
       ankle-width-right: 67.66
       ankle-width-units: millimeters
       gender: male
    trial:
       id: 58
       subject-id: 8
       datetime: 2014-03-28
       notes: >
           The subject did a somersault during this trial instead of following
           instructions to walk. Will have to use for another study.
       nominal-speed: 0.8
       nominal-speed-units: meters per second
       stationary-platform: True
       pitch: False
       sway: False
       hardware-settings:
           high-performance: True
       dflow-version: 3.16.1
       cortex-version: 3.1.1.1290
       marker-map:
           M1: LHEAD
           M2: THEAD
           M3: RHEAD
           M4: FHEAD
           M5: C7
       analog-channel-map:
           Channel1.Anlg: F1Y1
           Channel2.Anlg: F1Y2
           Channel3.Anlg: F1Y3
           Channel4.Anlg: F1X1
       events:
           A: Force Plate Zeroing
           B: Calibration Pose
           C: First Normal Walking
           D: Longitudinal Perturbation
           E: Second Normal Walking
           F: Unloaded End
       files:
           compensation: ../T057/mocap-057.txt
           mocap: mocap-058.txt
           record: record-058.txt
           meta: meta-058.yml
  \end{minted}
  \caption{A fictious example of a YAML formatted meta data file. All of the
    possible keys in the data set are shown.}
  \label{lis:example-meta-data}
\end{listing*}

\subsection*{Markers}
%
We make use of the full body 47 marker set described in \cite{Ton's HBM paper}
and presented in detail in Table \ref{tab:marker-labels}. As with all camera
based motion capture systems, the markers sometimes go missing in the
recording.

@@

    column indicates wheter the marker exists in the lower and/or full body

@@

    L/F & 7  & SACR  & Sacrum bone                           & On the scacral bone. \\

@@

\begin{enumerate}
  \item Loads the meta data file into a Python dictionary.
  \item Loads the DFlow mocap module TSV file into Pandas \verb|DataFrame|.
  \item Relabels the column headers to more meaningful names if this is
    specified in the meta data.
  \item Optionally identifies the missing values in the mocap marker data and
    replaces them with \verb|numpy.nan|.
  \item Optionally interpolates the missing marker values and replaces them
    with interpolated estimates using a variety of interpolation methods.
  \item Loads the D-Flow record module TSV file into a Pandas \verb|DataFrame|.
  \item Extracts the events and creates a dictionary mapping the event names in
    the meta data to the events detected in the record module file.
  \item Compensates the ground reaction loads based on whether the meta data
    indicates there was treadmill motion.
  \item Merges the data from the mocap module and record module into one data
    frame at the maximum common constant sample rate.
\end{enumerate}

@@

\begin{listing*}
  \begin{minted}[gobble=4]{python}
    >>> from gaitanalysis.gait import GaitData
    >>> gdata = GaitData(event_df)
    >>> gdata.inverse_dynamics_2d(left_markers, right_markers, left_loads,
    ...                           right_loads, mass, 6.0)
    >>> gdata.grf_landmarks('Right Fy', 'Left Fy', threshhold=20.0)
    >>> gdata.split_at('right')
    >>> gdata.plot_gait_cycles('Left Hip Joint Torque', mean=True)
    >>> gdata.save('gait-data.h5')
  \end{minted}
  \cprotect\caption{Python interpreter session showing how one could compute
    the inverse dynamics (joint angles and torques), indentify the gait
    landmarks (e.g. heelstrikes), split the data with respect to the gait
    landmarks in a Pandas \verb|Panel|, plot the mean and standard deviation of
    one time history with respect to the gait cycles, and save the data to disk.}
  \label{lis:dflowdata}
\end{listing*}

A similar workflow was used to produce Figure~\ref{fig:angle-torque-comparison}
which compares the mean and standard deviation of saggital plane joint angles
and torques from the perturbed gait cycles and the unperturbed gait cycles
computed from trial 20 for 120 gait cycles. The variation in the dynamics is
larger for the perturbed gait cycles.

@@

%{\small\bibliographystyle{unsrt}
%\bibliography{bibfile}}

>>>>

\author[1]{Jason K. Moore}
\author[1]{Sandra K. Hnat}
\author[1]{Antonie J. van den Bogert}

@@

  Here we share a rich gait data set collected from fifteen subjects walking at
  three speeds on an instrumented treadmill. Each trial consists of 120 seconds
  of normal walking and 480 seconds of walking while being longitudinally
  and/or laterally perturbed during each stance phase with pseudo-random
  fluctuations in the speed of the treadmill belt. A total of approximately 1.5
  hours of normal walking ($>5000$ gait cycles) and 6 hours of perturbed walking
  ($>20,000$ gait cycles) is included in the data set. We provide full body
  marker trajectories and ground reaction loads in addition to a presentation of
  processed data that includes gait landmarks, 2D joint angles, angular rates,
  and joint torques and the software used for the computations. The protocol is
  described in detail along with additional elaborate meta data. This data can
  likely be useful for validating or generating mathematical models that are
  capable of simulating normal periodic gait and non-periodic, perturbed gaits.

@@

\section*{Introduction}
%
The collection of dynamical data during human walking has a long history
beginning with the first motion pictures and now with modern marker based
motion capture techniques, image and video analysis, and high fidelity ground
reaction load measurements. Even though years of data on thousands of subjects
now exist, this data is not widely disseminated, well organized, nor available
with few or no restrictions. David Winter's published normative gait
data~\cite{Winter1990} is widely used in biomechanics studies, yet it comes
from relatively few subjects and only a small number of gait cycles per
subject. This small source has successfully inspired powered prosthetic control
design~\cite{GoldfarbXXXX}, but success in other fields using large sets of
data for discovery lead one to believe that may be possible in the
biomechanical field too. To find out if that hypothesis is true, our data needs
to be shared extensively, organized, and curated to enabled analysts.

There are some notable gait data sets and databases besides Winter's authoritative
set that are publicly available. Notably the International Society of
Biomechanics has maintained a web page (http://isbweb.org/data) since
approximately 1995 that includes data sets for download and most unencumbered
use. Vaughn, et. al's data~\cite{Vaughan1992} with kinematics and force plate
measurement from several subjects. Chris Kirtley shares normative gait from
several studies on the CGA Normative Gait Database~\cite{Kirtley} and these
files have influenced and been included in other studies, for
example the average gait cycles from children used in~\cite{Bogert2003}.

Chester et. al~\cite{Chester2007} report on a large gait database comparison
where one database contained kinematic data of 409 gait cycles of children
from 1 to 7 years old. Tirosh et. al, recognized the need of a comprehensive
data base for clinical gait data and created the Gaitabase~\cite{Tirosh2010}.
This database may contain a substantial amount of data but it is encumbered by
a very complicated and restrictive license and sharing scheme. The University
of Wisconsin at LaCrosse has a normative gait data set~\cite{Wilson} from 25
subjects with lower extremity marker data and force plate measurements from a
single stance phase.

More recent examples of biomechanist's sharing their data alongside
publications are: \cite{Bogert2013} which includes processed joint kinematics
from eleven subjects walking for few gait cycles and \cite{Wang2014} who
includes data from 10 subjects walking for five minutes each at five different
speeds but only a small set of lower extremity markers are present. The second
is notable because it publishes the data in Dryad, a modern citable data
repository.

The publicly available gait data is small compared to the number of gait
studies that have been performed over the years. And the data that is available
in general suffers from things such as few subjects, few gait cycles, few
markers, no raw data, limited force plate measurements, lack of meta data, and
restrictive licensing. To help with this situation we are making the data we
collected for our research purposes publicly available and free of the
previously mentioned deficiencies. Not only do we provide a larger set of
normative gait data that has been previously available we include an even large
set of data in which the subject is being perturbed which does not exist at
all. We believe both of these sets of data can serve a variety of use cases and
hope that we can save time for researchers with use cases for this data.

Our use case for the data is centered around emerging powered prosthetics and
orthotics need of bio-inspired control systems. Ideally, a powered prosthetic
would behave in such a way that the user would feel like their limb was never
disabled. There are a variety of approaches to developing bio-inspired control
systems some of which aim to mimic the reactions and motion of an able-bodied
person. A modern gait lab is able to collect a variety of kinematic, kinetic,
and physiological data from humans during gait. This data can potentially be
used to drive the design of the human-mimicking controller. With a rich enough
data set, one may be able to identify control mechanisms used during a human's
natural gait and recovery from perturbations. We have collected data that we
believe is richer than previous gait data sets and may be rich enough for
control identification.  The data can also be used for verification purposes
for controllers that have been designed in other manners.

With all of this in mind, we collected over five and half hours of gait data
from fifteen able bodied subjects and includes over 25,000 gait cycles. Most of
the subjects walked at three different speeds on an instrumented treadmill
while we collected full body marker locations and ground reaction loads from a
pair of force plates. The protocol for the majority of the trials includes two
minutes of normal walking and eight minutes of walking under the influence of
pseudo-random belt speed fluctuations. The data has been organized complete with
rich meta data and made available in the most unrestrictive form for other
research uses following modern best practices in data sharing~\cite{White2013}.

Furthermore, we include a small Apache licensed open source software library
for basic gait analysis and demonstrate it's use in the paper. The combination
of the open data and open software allow the results presented within to be
computationally reproducible and instructions are included in the associated
repository for doing so.

\section*{Participants}
%
Fifteen able bodied subjects including four females and eleven males with an
average age of $24\pm4$ years, height of $1.75\pm0.09$ m, mass of $74\pm13$ kg
participated in the study. The data has been anonymized with respect to the
participants' identities and an unique identification number was assigned to
each subject. A selection of the meta data collected for each subject is shown
in Table~\ref{tab:subjects}.
%
\begin{table*}
  \cprotect\caption{Information about the 15 participants. The final three
    columns give the trial numbers associated with each nominal treadmill
    speed. Additional trials with subject 0 are trials with no subject, i.e.
    unloaded trials that can be used for force compensation purposes, and are
    not shown in the table. Generated by \verb|src/subject_table.py|.}
  \centering
  \input{tables/subjects.tex}
  \label{tab:subjects}
\end{table*}

@@

\begin{itemize}
  \item A ForceLink R-Mill treadmill which has dual 6 degree of freedom force
    plates, independent belts for each foot, and motion capabilities (lateral,
    pitch).
  \item A 10 camera Motion Analysis motion capture system which includes Osprey
    cameras, the Cortex software (3.1.1.1290), and hardware for collecting
    analog and camera data simultaneously.
  \item National Instruments USB-6255 data acquisition unit.
  \item Four Sparkfun ADXL330 Triple Axis Accelerometer Breakout boards
    attached to the treadmill.
  \item Motek Medical’s D-Flow software (3.16.1 to 3.16.2) and visual display
    system.
\end{itemize}

@@

\section*{Protocol}
%
The experimental protocol consisted of static measurements and walking on the
treadmill for 10 minutes under unperturbed and perturbed conditions. Before a
set of trials on the same day the following happened:
%
\begin{itemize}
  \item Calibration of the motion capture system.
  \item Subject changes into athletic shoes, shorts, sports bra, baseball
    cap, and rock climbing harness.
  \item All 47 markers are applied directly to the skin except for the heel,
    toe, and head markers.~\footnote{The sacrum and rear pelvic markers were
    placed on the shorts for a small number of the subjects}.
  \item Subjects self-reported age, gender, and mass.
  \item Height is measured by the experimentalist.
  \item Four reference photographs (front, back, right, left) are taken of
    subjects marker locations.
\end{itemize}

After a briefing by the experimentalist on the trial protocol the subject
follows the verbal instructions of the experimentalist and the on-screen
instructions from the video display in front of the subject. The protocol goes
as so:
%
\begin{enumerate}
  \item Subject steps onto the treadmill and markers are identified with
    Cortex.
  \item The safety cable is attached loosely to the rock climbing harness such
    that no undue forces are acting on the subject during walking, but that the
    harness will prevent a full fall.
  \item The subject starts by stepping on sides of treadmill so that feet don't
    touch the force plates and the force plate signals are zeroed.
  \item Once notified by the screen, the subject stands in the initialization
    pose: standing straight up, looking forward, arms out by their sides (~45
    degrees) and the event is manually recorded by the operator.
  \item A countdown to the first normal walking phase is displayed. At the end
    of the countdown the treadmill starts at the specified speed and the
    subject is instructed to walk normally, to focus on the road in the
    screen, and not to look at their feet.
  \item After 1 minute of normal walking, the longitudinal perturbation phase
    begins.
  \item After 8 minutes of walking under the influence of the perturbations,
    the normal walking phase begins.
  \item After 1 minute, a countdown is shown and the treadmill decelerates to a
    stop.
  \item The subject is instructed to step off of the force plates for 10
    seconds.
  \item The subject could take a break before each additional trial.
\end{enumerate}

@@

The resulting output from the MATLAB script and Simulink produces a comma-
delimited text file of six columns: time stamp, slow, normal, and fast walking
perturbation signals, and slow and fast running signals.  These running signals
were generated in the event the testing protocol expanded from a walking study
to a running study, though they were not used in the data set.  The actual
motion of the treadmill belts are compared to the original input signals as
shown in Figure~\ref{fig:input_output}.   The standard deviations of the
outputs were calculated at 6.07\%, 9.68\%, and 12.68\% for the 0.8, 1.2,
and 1.6 $m \cdot s^{-1}$ speeds, respectively.

@@

The text file output from the mocap module in D-Flow is a tab delimited file.
The first line is the header and contains a time stamp column, frame number
column, marker position columns, force plate force/moment columns, force plate
center of pressure columns, other analog columns, and potentially results from
the real time Human Body Model \cite{Bogert2013} which is included with the
D-Flow software. These are detailed below. The numerical values of the
measurements are provided in decimal floating point notation with 6 decimals of
precision, e.g. 123456.123456 [\%1.6f].

@@

\begin{description}
  \item[TimeStamp] The monotonically increasing computer clock time when D-Flow
    receives a frame from Cortex. These are approximately at 100 hz and given
    in seconds.
  \item[FrameNumber] Monotonically increasing positive integers that correspond
    to each frame received from Cortex.
  \item[Marker Coordinates] Any column that ends in \verb+.PosX+, \verb+.PosY+,
    or \verb+.PosZ+ are marker coordinates expressed in Cortex's Cartesian
    reference frame. These values are in meters.
  \item[Ground Reaction Loads] There are three ground reaction forces and three
    ground reaction moments recorded by each of the two force plates in Newtons
    and Newton-Meters, respectively. The prefix for these columns is either
    \verb+FP1+ or \verb+FP2+ and represents either force plate 1 (left) or 2
    (right). The suffixes are either \verb+.For[XYZ]+, \verb+.Mom[XYZ]+ for the
    forces and moments, respectively. The force plate voltages are sampled at a
    much higher frequency than the cameras, but delivered at the Cortex camera
    sample rate, 100 Hz through the D-Flow mocap module. A force/moment
    calibration matrix stored in Cortex converts the voltages to forces and
    moments before sending it to D-Flow [1]. Cortex also computes the center of
    pressure from the forces, moments, and force plate dimensions. These have
    the same prefixes for the plate number, have the suffix \verb+.Cop[XYZ]+,
    and are in meters.
  \item[Analog Channels] Several analog signals are recorded under column
    headers \verb+Channel[1-99].Anlg+. These correspond to analog signals
    sampled by Cortex and correspond to the 96 analog-channel National
    Instruments 6255 DAQ box. The first twelve are the voltages from the
    force plate load cells. We also record the acceleration of 4 points on the
    treadmill base in channels 61-72.  TvdB: are these useful?  If so, we need to
    provide information on how to process these channels. If not, maybe
    not mention them at all?
\end{description}

@@

\begin{description}
  \item[analog-channel-map] A mapping of the strings D-Flow assigns to signals
    emitted from the analog channels of the NI USB-6255 to names the user
    desires.
  \item[cortex-version] The version of Cortex used to record the trial.
  \item[datetime]A date formatted string giving the date of the trial in the
    \verb|YYYY-MM-DD| format.
  \item[dflow-version] The version of D-Flow used to record the trial.
  \item[events] A key value map which prescribes names to the alphabetic events
    recorded in the record file.
  \item[files] A key value mapping of files associated with this trial
    where the key is the D-Flow file type and the value is the path to the file
    relative to the meta file.
  \item[hardware-settings] There are tons of settings for the hardware in both
    D-Flow, Cortex, and the other software in the system. This contains any
    non-default settings.
    \begin{description}
      \item[high-performance] A boolean value indicating whether the D-Flow high
        performance setting was on (True) or off (False).
    \end{description}
  \item[id] An unique three digit integer identifier for the trial. All of the
    file names and directories associated with this trial include this number.
  \item[marker-map] A key value map which maps marker names in the mocap file
    to the user's desired names for the markers.
  \item[marker-set] Indicates the HBM~\cite{Bogert2013} marker set used during
    the trial, either full, lower, or NA.
  \item[nominal-speed] A float representing the nominal desired treadmill speed
    during the trial.
  \item[nominal-speed-units] (string) The units of the nominal speed.
  \item[notes] Any notes about the trial.
  \item[pitch] A boolean that indicates if the treadmill pitch degree of
    freedom was actuated during the trial.
  \item[stationary-platform] A boolean that indicates whether the treadmill
      sway or pitch motion was actuated during the trial. If this flag is
      false, the measured ground reaction forces must be compensated for the
      inertial affects and be expressed in the motion capture reference frame.
  \item[subject-id] An integer corresponding to the subject in the trial.
  \item[sway] A boolean that indicates if the treadmill lateral (sway) degree
    of freedom was actuated during the trial.
\end{description}

\begin{listing*}
  \small
  \begin{minted}[gobble=4]{yaml}
    study:
       id: 1
       name: Gait Control Identification
       description: Perturb the subject during walking and running.
    subject:
       id: 8
       age: 20
       mass: 70.0
       mass-units: kilograms
       height: 1.572
       height-units: meters
       knee-width-left: 107.43
       knee-width-right: 107.41
       knee-width-units: millimeters
       ankle-width-left: 70.52
       ankle-width-right: 67.66
       ankle-width-units: millimeters
       gender: male
    trial:
       id: 58
       subject-id: 8
       datetime: 2014-03-28
       notes: >
           The subject did a somersault during this trial instead of following
           instructions to walk. Will have to use for another study.
       nominal-speed: 0.8
       nominal-speed-units: meters per second
       stationary-platform: True
       pitch: False
       sway: False
       hardware-settings:
           high-performance: True
       dflow-version: 3.16.1
       cortex-version: 3.1.1.1290
       marker-map:
           M1: LHEAD
           M2: THEAD
           M3: RHEAD
           M4: FHEAD
           M5: C7
       analog-channel-map:
           Channel1.Anlg: F1Y1
           Channel2.Anlg: F1Y2
           Channel3.Anlg: F1Y3
           Channel4.Anlg: F1X1
       events:
           A: Force Plate Zeroing
           B: Calibration Pose
           C: First Normal Walking
           D: Longitudinal Perturbation
           E: Second Normal Walking
           F: Unloaded End
       files:
           compensation: ../T057/mocap-057.txt
           mocap: mocap-058.txt
           record: record-058.txt
           meta: meta-058.yml
  \end{minted}
  \caption{A fictitious example of a YAML formatted meta data file. All of the
    possible keys in the data set are shown.}
  \label{lis:example-meta-data}
\end{listing*}

\subsection*{Markers}
%
We make use of the full body 47 marker set described in \cite{Bogert2013}
and presented in detail in Table \ref{tab:marker-labels}. As with all camera
based motion capture systems, the markers sometimes go missing in the
recording.

@@

    column indicates whether the marker exists in the lower and/or full body

@@

    L/F & 7  & SACR  & Sacrum bone                           & On the sacral bone. \\

@@

\begin{enumerate}
  \item Loads the meta data file into a Python dictionary.
  \item Loads the D-Flow mocap module TSV file into Pandas \verb|DataFrame|.
  \item Relabels the column headers to more meaningful names if this is
    specified in the meta data.
  \item Optionally identifies the missing values in the mocap marker data and
    replaces them with \verb|numpy.nan|.
  \item Optionally interpolates the missing marker values and replaces them
    with interpolated estimates using a variety of interpolation methods.
  \item Loads the D-Flow record module TSV file into a Pandas \verb|DataFrame|.
  \item Extracts the events and creates a dictionary mapping the event names in
    the meta data to the events detected in the record module file.
  \item Compensates the ground reaction loads based on whether the meta data
    indicates there was treadmill motion.
  \item Merges the data from the mocap module and record module into one data
    frame at the maximum common constant sample rate.
\end{enumerate}

@@

\begin{listing*}
  \begin{minted}[gobble=4]{python}
    >>> from gaitanalysis.gait import GaitData
    >>> gdata = GaitData(event_df)
    >>> gdata.inverse_dynamics_2d(left_markers, right_markers, left_loads,
    ...                           right_loads, mass, 6.0)
    >>> gdata.grf_landmarks('Right Fy', 'Left Fy', threshhold=20.0)
    >>> gdata.split_at('right')
    >>> gdata.plot_gait_cycles('Left Hip Joint Torque', mean=True)
    >>> gdata.save('gait-data.h5')
  \end{minted}
  \cprotect\caption{Python interpreter session showing how one could compute
    the inverse dynamics (joint angles and torques), identify the gait
    landmarks (e.g. heel strikes), split the data with respect to the gait
    landmarks in a Pandas \verb|Panel|, plot the mean and standard deviation of
    one time history with respect to the gait cycles, and save the data to disk.}
  \label{lis:dflowdata}
\end{listing*}

A similar work flow was used to produce Figure~\ref{fig:angle-torque-comparison}
which compares the mean and standard deviation of saggital plane joint angles
and torques from the perturbed gait cycles and the unperturbed gait cycles
computed from trial 20 for 120 gait cycles. The variation in the dynamics is
larger for the perturbed gait cycles.

@@

{\small\bibliographystyle{unsrt}
\bibliography{gait-data}}
